Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-07-01 submission/manuscript.Rmd
shigeki results edits and ChIPseq line count

In this study, we characterized the misexpression of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Previous work has demonstrated constitutive _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016]. Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus surprisingly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been systematically evaluated.

@@

To elucidate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We identified widespread dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, and ovary genes. The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are unique to germ cells. To better characterize germline gene misexpression, we then generated a dataset of germline-enriched genes by comparing gene expression in gonads with germ cell depletion. We found _Kdm5c_-KO EpiLCs primarily expressed unique germline genes compared to the mature _Kdm5c_-KO brain, including _Dazl_ and _Stra8_, key drivers of germline identity and meiosis. While KDM5C is directly bound to some germline gene promoters in EpiLCs, it is not directly bound to many germline-enriched mRNAs expressed with _Kdm5c_-KO cells, indicating germline genes can be aberrantly transcribed through indirect mechanisms. Finally, we found KDM5C loss impairs the placement of DNA methylation at germline gene promoters as ESCs differentiate into EpiLCs. Therefore, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including establishment of the soma-germline boundary.

@@

A previous RNA sequencing (RNA-seq) study surprisingly revealed the aberrant expression of testis genes within the adult _Kdm5c_ knockout (-KO) hippocampus[@scandagliaLossKdm5cCauses2017]. Given the high abundance of testis-enriched genes within the mouse transcriptome, it is currently unclear if their misexpression in the _Kdm5c_-KO brain is indicative of random de-repression throughout the genome or instead reflects an impairment in tissue identity. To distinguish these possibilities, we globally assessed the expression of previously characterized tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our previously published mRNA-seq datasets of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020]

@@

To further validate if these testis DEGs are truly germline genes, we then compared their expression in a published RNA-seq dataset of wild-type (WT) and germ cell-depleted (Kit^W/Wv^) mouse testes [@muellerIndependentSpecializationHuman2013]. We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 2B). The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012]. We additionally assessed testis DEG expression in a published testis single cell RNA-seq dataset that identified cell type-specific markers[@greenComprehensiveRoadmapMurine2018]. We found that while some testis-enriched DEGs were classified as specific markers for different germ cell types (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids) none marked somatic cells (Figure 2C). Together, these data suggest the _Kdm5c_-KO brain expresses germline genes.

We then aimed to globally characterize germline gene misexpression in _Kdm5c_-KO mice, but lacked a comprehensive list of mouse germline-enriched genes. To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in wild-type and germline-depleted (Kit^W/Wv^) mice. We assessed gene expression in male and female mice at embryonic day 12, 14, and 16, as well as male postnatal day 6 and adult testes. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes.

@@

We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom) and found no significant difference in primed pluripotency genes that indicate proper EpiLC differentiation (Figure 3B). We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and again observed widespread dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3C).

Using our curated list, we found `r format(nrow(epigerm))` germline-enriched genes were misexpressed in _Kdm5c_-KO EpiLCs, including _Cyct_ (Figure 3D). To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across sequencing datasets (Figure 3E).

We then compared the function of EpiLC and brain germline DEGs through gene ontology and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3F), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). While a few meiosis-related genes are misexpressed in the _Kdm5c_-KO brain, DEGs unique to _Kdm5c_-KO EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo. However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`).

_Dazl_ is an RNA-binding protein that regulates the translation of germline mRNAs and is essential for germ cell function[@liDAZLMasterTranslational2019]. Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_-KO cellular function. We therefore stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H). We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020]. Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including master regulators of germline identity that can be translated into protein.

@@

Previous work suggests KDM5C represses germline genes during early development, since KDM5C binding is gradually lost at hippocampal germline DEG promoters over development[@scandagliaLossKdm5cCauses2017] and re-expression of KDM5C in neuronal cultures fails to suppress their transcription. However, we identified many germline genes that were only dysregulated in _Kdm5c_-KO EpiLCs and not in the mature brain, including master regulators of meiosis and germline identity. While KDM5C was recently shown to bind to _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023], it's unclear if KDM5C binds to other master regulators that are expressed in _Kdm5c_-KO EpiLCs and if this binding is maintained as cells differentiate.

To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets we previously generated in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary hippocampal and cortical neuron cultures (PNCs)[@iwaseMouseModelXlinked2016]. We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: 2,437, PNCs: 1,610, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: , PNCs: , +/- kb TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ) (Figure 4A). Gene ontology analysis of KDM5C-bound promoters in EpiLCs revealed high enrichment of germline-related ontologies, including meiotic nuclear division (GO:0140013) and synaptonemal complex assembly (GO:0007130) (Figure 4B). Contrastingly, germline ontologies are not enriched in PNCs and KDM5C-bound genes instead perform non-germline-specific functions, such as heterocycle synthesis (GO: 0018130) and pyrimidine metabolism (GO: 0072527).

@@

# Discussion
- The demarcation of the germ vs soma is a key feature of multicellularity
- Other H3K4me regulators, anything known about tissue-biased gene expression?
- tissue-biased gene expression:
	- However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function. For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory (Figure 1E).
- Otx2 is properly expressed in EpiLCs and prevents pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/

@@

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
  \caption[\textit{Kdm5c}-KO epiblast-like cells express master regulators of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express master regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells. Middle - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs). Bottom - representative images of wild-type (WT) and \textit{Kdm5c}-KO ESC to EpiLC differentiation. Brightfield images taken at 20X. \textbf{B.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs. Welch's t-test, expression in transcripts per million (TPM). \textbf{C.} Number of tissue-enriched differentially expressed genes (DEGs). * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test. \textbf{D.} Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs. \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets. \textbf{F.} enrichPlot comparing enriched gene ontologies for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs. \textbf{G.} Top - Example germline identity DEGs unique to EpiLCs, p-values for Welch's t-test. Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs \textbf{H.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs. Percentage of DAZL-positive cells normalized to DAPI, p-value for Welch's t-test}
  \label{figurelabel}
\end{figure}

@@

**Figure 3: Kdm5c-KO epiblast-like cells express master regulators of germline identity**
	* A) ESC to EpiLC differentiation Left - Morphology is unchanged, 
	* B) 5CKO EpiLCs express EpiLC differentiation genes similar to WT lvls
	* C) Male EpiLCs express germline genes (example Cyct again)
	* Overlap between brain and EpiLC germline genes - show they're mostly unique
	* GO of Brain and EpiLC germline genes (meiotic enriched) 
	* Bigwigs or TPM of master regulators
		* Show that while some are also 2-cell genes (Dazl), 2-cell specific genes aren't dysregulated (Zscan4). Important point because published KDM5C dazl paper is saying KDM5C is a 2-cell regulator, but as far as I can tell only genes shared between germline and 2-cell are dysregulated.  
	* Staining of Dazl (+ Stra8 if I can get it to work)

>>>>

In this study, we characterized the misexpression of tissue-enriched genes with loss of the chromatin regulator lysine demethylase 5C (KDM5C, also known as SMCX or JARID1C), a histone 3 lysine 4 demethylase. Pathogenic mutations in _KDM5C_ cause Intellectual Developmental Disorder, X-linked, Syndromic, Claes-Jensen Type (MRXSCJ, OMIM: 300534), whose features include short stature, intellectual disability, seizures, aberrant aggression, and autistic behaviors[@claesNovelSyndromicForm2000; @jensenMutationsJARID1CGene2005; @carmignacFurtherDelineationFemale2020]. Previous work has demonstrated constitutive _Kdm5c_ knockout (-KO) mice recapitulate key MRXSCJ patient phenotypes, including hyperaggression and learning impairments[@iwaseMouseModelXlinked2016]. Next generation RNA sequencing (RNA-seq) in the _Kdm5c_-KO hippocampus unexpectedly revealed ectopic expression of testis-enriched genes within the brain[@scandagliaLossKdm5cCauses2017]. However, it is currently unclear if misexpression in the _Kdm5c_-KO brain is unique to testis genes, as other tissue-enriched genes have not been systematically evaluated.

@@

To elucidate KDM5C's role in tissue identity, here we characterized the aberrant transcription of tissue-enriched genes within the _Kdm5c_-KO brain and epiblast-like cells (EpiLCs), an _in vitro_ model of the post-implantation embryo. We identified general dysregulation of tissue-enriched genes in both the adult _Kdm5c_-KO brain and EpiLCs, including misexpression of liver, muscle, and ovary genes. The _Kdm5c_-KO amygdala and hippocampus had significant enrichment of testis-biased genes that are unique to germ cells. To better characterize germline gene misexpression, we then generated a dataset of germline-enriched genes by comparing gene expression in gonads with germ cell depletion. We found _Kdm5c_-KO EpiLCs primarily expressed unique germline genes compared to the mature _Kdm5c_-KO brain, including _Dazl_ and _Stra8_, key drivers of germline identity and meiosis. While KDM5C is directly bound to some germline gene promoters in EpiLCs, it is not directly bound to many germline-enriched mRNAs expressed with _Kdm5c_-KO cells, indicating germline genes can be aberrantly transcribed through indirect mechanisms. Finally, we found KDM5C loss impairs the placement of DNA methylation at germline gene promoters as ESCs differentiate into EpiLCs. Therefore, we propose KDM5C plays a crucial role in the development of tissue identity during early embryogenesis, including establishment of the soma-germline boundary.

@@

Previous RNA sequencing (RNA-seq) in the adult _Kdm5c_ knockout (-KO) hippocampus identified a few transcripts that are typically expressed in the testis[@scandagliaLossKdm5cCauses2017]. However, the expression of tissue-specific genes within the _Kdm5c_-KO brain has not systematically characterized. Currently, it is unclear if other tissue-specific transcripts are also expressed within the _Kdm5c_-KO brain and if misexpression of testis-enriched genes are a major consequence of KDM5C loss, given the high abundance of testis-enriched genes within the mouse transcriptome. Therefore, we first globally assessed the expression of previously characterized tissue-enriched genes[@liComprehensiveMouseTranscriptomic2017] in our published mRNA-seq datasets of the _Kdm5c_-KO amygdala and hippocampus[@vallianatosMutuallySuppressiveRoles2020]

**note: the idea is there but need to work on wording. The extent to which brain tissue identity is altered is unknown and if testis genes are a major effect**

@@

To further validate if these testis DEGs are truly germline genes, we then evaluated their expression in germ cell-depleted testes via  _c-Kit_ mutation (Kit^W/Wv^)[@muellerIndependentSpecializationHuman2013]. Heterozygous _W_ and _Wv_ mutations in the enzymatic domain of _c-Kit_ prevents the maturation of germ cells, resulting in overall germ cell depletion[@handelSertoliCellDifferentiation1979]. We found almost all _Kdm5c_-KO testis-enriched DEGs lose their expression with germ cell depletion (Figure 2B). The only testis DEG that did not show considerable downregulation with germline depletion was _FK506 binding protein 6 (Fkbp6)_, a known regulator of piRNA expression and meiosis in germ cells[@crackowerEssentialRoleFkbp62003; @xiolRoleFkbp6Chaperone2012]. We additionally assessed testis DEG expression in a published testis single cell RNA-seq dataset that identified cell type-specific markers[@greenComprehensiveRoadmapMurine2018]. We found that while some testis-enriched DEGs were classified as specific markers for different germ cell types (e.g. spermatogonia, spermatocytes, round spermatids, and elongating spermatids) none marked somatic cells (Figure 2C). Together, these data suggest the _Kdm5c_-KO brain expresses germline genes.

We then aimed to globally characterize germline gene misexpression in _Kdm5c_-KO mice, but lacked a comprehensive list of mouse germline-enriched genes. To facilitate downstream analyses, we generated a list of male and female germline-enriched genes by evaluating gene expression in wild-type and germline-depleted (Kit^W/Wv^) mice. We assessed gene expression in male and female mice at embryonic day 12, 14, and 16, as well as male postnatal day 6 and adult testes. We defined genes as germline-enriched if their expression met the following criteria: 1) their expression is greater than 1 FPKM in wild-type germline 2) their expression in any wild-type, non-gonadal tissue[@liComprehensiveMouseTranscriptomic2017] does not exceed 20% of maximum expression in wild-type germline, and 3) their expression in the germ cell-depleted germline, for any sex or time point, does not exceed 20% of their maximum expression in the wild-type germline. These criteria yielded `r nrow(germgenes)` germline-enriched genes (Figure 2D), which was hereafter used as a resource for assessing misexpression of germline genes (Supplementary table 1).

@@

We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom) and found no significant difference in primed pluripotency genes that indicate proper EpiLC differentiation (Figure 3B). We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and again observed general dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis, although they were not significantly enriched (Figure 3C).

Using our curated list, we found `r format(nrow(epigerm))` germline-enriched genes were misexpressed in _Kdm5c_-KO EpiLCs, including _Cyct_ (Figure 3D). To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we then compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. The majority of EpiLC germline DEGs were unique to EpiLCs, with only _Cyct_ shared across sequencing datasets (Figure 3E). We then compared the function of EpiLC and brain germline DEGs through gene ontology and found particularly high enrichment of meiosis-related gene ontologies in EpiLCs (Figure 3F), such as meiotic cell cycle (`r getGO("meiotic cell cycle", GO_EpiBrain, 'ID', "EpiLC")`, p.adjust = `r signif(getGO("meiotic cell cycle", GO_EpiBrain, 'p.adjust', "EpiLC"), 3)`). Contrastingly, the _Kdm5c_-KO hippocampus instead expressed genes involved in late-stage sperm development, those involved in the sperm axoneme. This suggests the germline developmental program may be occuring ectopically as _Kdm5c_-KO cells progresses through somatic tissue development **(note: placement is awkward. Even put here or talk about in dicussion because ChIPseq could help support this? - KDM5C not bound to a lot of brain DEGs)**.

While a few meiosis-related genes are misexpressed in the _Kdm5c_-KO brain, DEGs unique to _Kdm5c_-KO EpiLCs included key drivers of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019]. Of note, some germline genes, including _Dazl_, are also expressed in the two-cell embryo. However, we did not see misexpression of two-cell embryo-specific genes, like _Zscan4d_ (q = `r signif(zscan4d, 3)`), indicating _Kdm5c_-KO in EpiLCs does not revert cells back to a 2-cell state (Figure 3G).

_Dazl_ is an RNA-binding protein that regulates the translation of germline mRNAs and is essential for germ cell function[@liDAZLMasterTranslational2019]. Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_-KO cellular function. We therefore stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H). We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020]. Altogether these results suggest _Kdm5c_-KO EpiLCs fail to decommission germline genes during early embryogenesis, including key drivers of germline identity that can be translated into protein.

@@

Previous work suggests KDM5C represses germline genes during early development, since KDM5C binding is gradually lost at hippocampal germline DEG promoters over development[@scandagliaLossKdm5cCauses2017] and re-expression of KDM5C in neuronal cultures fails to suppress their transcription. However, we identified many germline genes that were only dysregulated in _Kdm5c_-KO EpiLCs and not in the mature brain, including key regulators of meiosis and germline identity. While KDM5C was recently shown to bind to _Dazl_ in ESCs[@guptaGenomewideScreenReveals2023; @aladhamiSystematicIdentificationFactors2023], it's unclear if KDM5C binds to other germline drivers that are expressed in _Kdm5c_-KO EpiLCs and if this binding is maintained as cells differentiate.

```{r, KDM5C_ChIPseq, results='hide', echo = FALSE, warning=FALSE,message=FALSE,error=FALSE}
KDM5C_ChIP_EpiLC <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_EpiLC.csv", header = T)

KDM5C_ChIP_PNC <- read.csv("../data/processed/KDM5C_ChIPseq_boundpromoters_PNC.csv", header = T)

To better elucidate the mechanism behind KDM5C-mediated germline gene silencing, we analyzed KDM5C chromatin immunoprecipitation followed by DNA sequencing (ChIP-seq) datasets we previously generated in EpiLCs[@samantaActivationXistEvolutionarily2022] and primary hippocampal and cortical neuron cultures (PNCs)[@iwaseMouseModelXlinked2016]. We identified a higher number of KDM5C peaks in EpiLCs overall (EpiLCs: `r nrow(KDM5C_ChIP_EpiLC)`, PNCs: `r nrow(KDM5C_ChIP_PNC)`, MACS2 q < 0.1 and fold enrichment > 1) and while KDM5C was primarily localized to gene promoters in both cell types (EpiLCs: , PNCs: , +/- kb TSS), PNCs showed increased localization to distal intergenic regions (EpiLCs: , PNCs: ) (Figure 4A). Gene ontology analysis of KDM5C-bound promoters in EpiLCs revealed high enrichment of germline-related ontologies, including meiotic nuclear division (GO:0140013) and synaptonemal complex assembly (GO:0007130) (Figure 4B). Contrastingly, germline ontologies are not enriched in PNCs and KDM5C-bound genes instead perform non-germline-specific functions, such as heterocycle synthesis (GO: 0018130) and pyrimidine metabolism (GO: 0072527).

@@

# Discussion
- The demarcation of the germ vs soma is a key feature of multicellularity
- Other H3K4me regulators, anything known about tissue-biased gene expression?
- This suggests the germline developmental program is occuring ectopically as _Kdm5c_-KOs progresses through somatic tissue development
- tissue-biased gene expression:
	- However unlike the gonadal-biased DEGs, many liver and muscle-biased DEGs have a known involvement in brain function. For example, the liver-biased DEG is _Apolipoprotein C-I (Apoc1)_, is important for lipoprotein metabolism but has also been shown to influence learning and memory (Figure 1E).
- Otx2 is properly expressed in EpiLCs and prevents pgc identity https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6485399/

@@

\begin{figure}
  \centering 
  \includegraphics{../submission/compiled_figs/EpiLCs_too.pdf}
  \caption[\textit{Kdm5c}-KO epiblast-like cells express key drivers of germline identity ]{\textbf{\textit{Kdm5c}-KO epiblast-like cells express key drivers regulators of germline identity A.} Top - Diagram of \textit{in vivo} differentiation of embryonic stem cells (ESCs) of the inner cell mass into epiblast stem cells. Middle - \textit{in vitro} differentiation of ESCs into epiblast-like cells (EpiLCs). Bottom - representative images of wild-type (WT) and \textit{Kdm5c}-KO ESC to EpiLC differentiation. Brightfield images taken at 20X. \textbf{B.} No significant difference in primed pluripotency maker expression in wild-type versus \textit{Kdm5c}-KO EpiLCs. Welch's t-test, expression in transcripts per million (TPM). \textbf{C.} Number of tissue-enriched differentially expressed genes (DEGs). * p<0.05, ** p<0.01, *** p<0.001, Fisher's exact test. \textbf{D.} Average bigwigs of an example germline gene, \textit{Cyct}, that is dysregulated \textit{Kdm5c}-KO EpiLCs. \textbf{E.} Upset plot displaying the overlap of germline DEGs expressed in \textit{Kdm5c}-KO EpiLCs, amygdala (AMY), and hippocampus (HIP) RNA-seq datasets. \textbf{F.} enrichPlot comparing enriched gene ontologies for \textit{Kdm5c}-KO EpiLC, amygdala, and hippocampus germline DEGs. \textbf{G.} Top - Example germline identity DEGs unique to EpiLCs, p-values for Welch's t-test. Bottom - Average bigwigs of \textit{Dazl} and \textit{Stra8} expression in wild-type and \textit{Kdm5c}-KO EpiLCs \textbf{H.} Immunocytochemistry of DAZL in male wild-type (WT) and \textit{Kdm5c}-KO (5CKO) EpiLCs. Percentage of DAZL-positive cells normalized to DAPI, p-value for Welch's t-test}
  \label{figurelabel}
\end{figure}

@@

**Figure 3: Kdm5c-KO epiblast-like cells express key drivers regulators of germline identity**
	* A) ESC to EpiLC differentiation Left - Morphology is unchanged, 
	* B) 5CKO EpiLCs express EpiLC differentiation genes similar to WT lvls
	* C) Male EpiLCs express germline genes (example Cyct again)
	* Overlap between brain and EpiLC germline genes - show they're mostly unique
	* GO of Brain and EpiLC germline genes (meiotic enriched) 
	* Bigwigs or TPM of master regulators
		* Show that while some are also 2-cell genes (Dazl), 2-cell specific genes aren't dysregulated (Zscan4). Important point because published KDM5C dazl paper is saying KDM5C is a 2-cell regulator, but as far as I can tell only genes shared between germline and 2-cell are dysregulated.  
	* Staining of Dazl (+ Stra8 if I can get it to work)
